Dynamics of Laterally Propagating Flames in X-ray Bursts. I. Burning Front Structure

### 2019-12-09 xrb/paper.tex
keywords

\keywords{convection---hydrodynamics---methods: numerical---stars: neutron---X-rays: bursts}

@@

\begin{figure}[t]
	\plotone{grids.pdf}
	\caption{\label{fig:grids}A section of a 2D simulation showing the four-level
		grid structure. Note that the boxes shown are not the simulation zones, which span
		$10~\mathrm{cm}$ at the finest level, but subdomains containing approximately equal
        numbers of zones that are distributed across MPI processes. A coarse base grid
		extending to the upper boundary is drawn in white, with the magenta, green, and
		blue grids showing the jumps in refinement needed to fully resolve the fuel layer
		and underlying neutron star.}
\end{figure}

@@

\subsection{Flame Propagation}

To measure the propagation rate of the burning front, we first collapse our nuclear
energy generation rate ($\dot{e}_\mathrm{nuc}$) data at each time into a 1D radial
profile by averaging over the vertical coordinate. We then take the peak
$\dot{e}_\mathrm{nuc}$ value across all profiles to provide a fixed reference point.
We define the position of the flame front to be the location ahead of hottest part of
the flame where $\dot{e}_\mathrm{nuc}$ first drops to $< 0.1 \%$ of the global maximum. 
This corresponds roughly to the leading edge of the burning region. Averaging
over the vertical coordinate helps to reduce sensitivity to localized fluid motions, as does
tracking the $0.1 \%$ contour rather than a local maximum. As we see in Figure
\ref{fig:flame_speed}, the flame settles into a state of steady propagation after an initial
transient period spanning $\sim 3$ ms. The position data here are well fitted by a linear
function of time, and the resulting slope gives the velocity of the flame front.

@@

All of the various approximations have the expected effects. The flame in the 5/5
boost run propagates at about half the speed of the 10/10 flame. This is consistent
with the 1D laminar flames, and is the behavior predicted by Eq.~\ref{eq:flame_scaling}. The
1000~Hz run goes 2 times faster than the 10/10 2000~Hz run, as expected from the
inverse relation between rotation rate and the ratio of Rossby length to scale height \citep{cavecchi:2013}.
This confirms that the role of the Coriolis
force is to limit the rate of flame spreading by the anticipated geometrical/hydrodynamical
effect ($s_\mathrm{front} = s_L * L_R / H$). Reducing the resolution had minimal impact on
the flame speed -- the low resolution line is right on top of the fit for the standard run
in Figure \ref{fig:flame_speed}, with their slopes differing by only a few percent. Using a
smaller network also produces similar behavior to the standard run, although there is a small
reduction in speed owing to less energetic burning.

>>>>

\keywords{X-ray bursts (1814), Nucleosynthesis (1131), Hydrodynamical simulations (767), Hydrodynamics (1963), Neutron stars (1108), Open source software (1866), Computational methods (1965)}

@@

\begin{figure}[t]
	\plotone{grids.pdf}
	\caption{\label{fig:grids}A section of a 2D simulation showing the four-level
		grid structure. Note that the boxes shown are not the simulation zones, which span
		$10~\mathrm{cm}$ at the finest level, but subdomains containing approximately equal
        numbers of zones that are distributed across MPI processes. A coarse base grid
		extending to the upper boundary is drawn in white, with the green, blue, and black
		grids showing the jumps in refinement needed to fully resolve the fuel layer and
		underlying neutron star.}
\end{figure}

@@

\subsection{Flame Acceleration}

To measure the propagation \MarginPar{Can we rename this section "Flame Propagation"? The burning front in our simulations seems to move at a constant velocity, and we don't discuss acceleration.}
rate of the burning front, we first collapse our nuclear
energy generation rate ($\dot{e}_\mathrm{nuc}$) data at each time into a 1D radial
profile by averaging over the vertical coordinate. We then take the peak
$\dot{e}_\mathrm{nuc}$ value across all profiles to provide a fixed reference point.
We define the position of the flame front to be the location ahead of hottest part of
the flame where $\dot{e}_\mathrm{nuc}$ first drops to $< 0.1 \%$ of the global maximum. 
This corresponds roughly to the leading edge of the burning region. Averaging
over the vertical coordinate helps to reduce sensitivity to localized fluid motions, as does
tracking the $0.1 \%$ contour rather than a local maximum. As we see in Figure
\ref{fig:flame_speed}, the flame settles into a state of steady propagation after an initial
transient period spanning $\sim 3$ ms. The position data here are well fitted by a linear
function of time, and the resulting slope gives the velocity of the flame front.

@@

All of the various approximations have the expected effects. The flame in the 5/5
boost run propagates at about half the speed of the 10/10 flame. This is consistent
with the 1D laminar flames, and is the behavior predicted by Eq.~\ref{eq:flame_scaling}. The
1000~Hz run goes 2 times faster than the 10/10 2000~Hz run, as expected from the
inverse relation between rotation rate and the ratio of the burning front area to vertical area \citep{cavecchi:2013}.
This confirms that the role of the Coriolis
force is to limit the rate of flame spreading by the anticipated geometrical/hydrodynamical
effect ($s_\mathrm{front} = s_L * L_R / H$). Reducing the resolution had minimal impact on
the flame speed -- the low resolution line is right on top of the fit for the standard run
in Figure \ref{fig:flame_speed}, with their slopes differing by only a few percent. Using a
smaller network also produces similar behavior to the standard run, although there is a small
reduction in speed owing to less energetic burning.
